---
title: "MDTS4214_737_PROJECT"
author: "SULAGNA ROY"
date: "2026-03-28"
output: pdf_document
---
# The Data Set:
```{r}
data=read.csv("Students_Academic_Performance.csv")
data=data[,2:ncol(data)]
```

# The Dimension of the Data Set:

## No. of Students:
```{r}
n=nrow(data);n
```
## No. of Variables:
```{r}
p=ncol(data);p
```
## Knowing the variables:
```{r}
colnames(data)
```
```{r}
attach(data)
```

```{r}
table(gender)
```
```{r}
table(grade_level)
```
```{r}
table(ai_tools_used)
```
```{r}
table(ai_usage_purpose)
```
```{r}
table(ai_dependency_score)
```
```{r}
table(ai_ethics_score)
```
```{r}
table(concept_understanding_score)
```
```{r}
table(study_consistency_index)
```
```{r}
table(improvement_rate)
```
```{r}
table(class_participation_score)
```
## Coding Of Categorical Variables:
### 1.Gender
```{r}
gender
```
```{r}
Gender=matrix(0,ncol=2,nrow=n)
colnames(Gender)=c("F","M")
for(i in 1:n)
{
  if(gender[i]=="Female")
  {
    Gender[i,]=c(1,0)
  }else if(gender[i]=="Male")
  {
    Gender[i,]=c(1,1)
  }else
  {
    Gender[i,]=c(0,0)
  }
}
CD1=data.frame(gender,Gender);CD1
```
### 2.Grade_Level
```{r}
grade_level
```
```{r}
Grade_Level=matrix(0,ncol=5,nrow=n)
colnames(Grade_Level)=c("grade_level_11","grade_level_12","grade_level_13","grade_level_14","grade_level_15")
for(i in 1:n)
{
  if(grade_level[i]=="10th")
  {
    Grade_Level[i,]=c(0,0,0,0,0)
  }else
  {
    if(grade_level[i]=="11th")
    {
      Grade_Level[i,]=c(1,0,0,0,0)
    }else
    {
      if(grade_level[i]=="12th")
      {
        Grade_Level[i,]=c(1,1,0,0,0)
      }else if(grade_level[i]=="1st Year")
      {
        Grade_Level[i,]=c(1,1,1,0,0)
      }else
      {
        if(grade_level[i]=="2nd Year")
        {
          Grade_Level[i,]=c(1,1,1,1,0)
        }else
        {
          Grade_Level[i,]=c(1,1,1,1,1)
        }
      }
    }
  }
}
CD2=data.frame(grade_level,Grade_Level);CD2
```
### 3.AI tools Used
```{r}
ai_tools_used
```


```{r}
AI_Tools_Used=matrix(0,ncol=5,nrow=n)
colnames(AI_Tools_Used)=c("ai_tool_1", "ai_tool_2", "ai_tool_3", "ai_tool_4", "ai_tool_5")
for(i in 1:n)
{
  if(ai_tools_used[i]=="ChatGPT")
  {
    AI_Tools_Used[i,]=c(0,0,0,0,0)
  }else
  {
    if(ai_tools_used[i]=="ChatGPT+Gemini")
    {
      AI_Tools_Used[i,]=c(1,0,0,0,0)
    }else
    {
      if(ai_tools_used[i]=="Claude")
      {
        AI_Tools_Used[i,]=c(1,1,0,0,0)
      }else if(ai_tools_used[i]=="Copilot")
      {
        AI_Tools_Used[i,]=c(1,1,1,0,0)
      }else
      {
        if(ai_tools_used[i]=="Gemini")
        {
          AI_Tools_Used[i,]=c(1,1,1,1,0)
        }else
        {
          AI_Tools_Used[i,]=c(1,1,1,1,1)
        }
      }
    }
  }
}
CD3=data.frame(ai_tools_used,AI_Tools_Used);CD3
```
### 4.AI Usage Purpose
```{r}
ai_usage_purpose
```

```{r}
AI_Usage_Purpose=matrix(0,ncol=5,nrow=n)
colnames(AI_Usage_Purpose)=c("Code","Doubt","Exam","HW","Note")
for(i in 1:n)
{
  if(ai_usage_purpose[i]=="None")
  {
    AI_Usage_Purpose[i,]=c(0,0,0,0,0)
  }else
  {
    if(ai_usage_purpose[i]=="Coding")
    {
      AI_Usage_Purpose[i,]=c(1,0,0,0,0)
    }else
    {
      if(ai_usage_purpose[i]=="Doubt Solving")
      {
        AI_Usage_Purpose[i,]=c(1,1,0,0,0)
      }else if(ai_usage_purpose[i]=="Exam Prep")
      {
        AI_Usage_Purpose[i,]=c(1,1,1,0,0)
      }else
      {
        if(ai_usage_purpose[i]=="Homework")
        {
          AI_Usage_Purpose[i,]=c(1,1,1,1,0)
        }else
        {
          AI_Usage_Purpose[i,]=c(1,1,1,1,1)
        }
      }
    }
  }
}
CD4=data.frame(ai_usage_purpose,AI_Usage_Purpose);CD4
```
### 5.AI Dependency Score
```{r}
ai_dependency_score
```
```{r}
AI_Dependency_Score=matrix(0,ncol=9,nrow=n)
colnames(AI_Dependency_Score)=c("AIDS2","AIDS3","AIDS4","AIDS5","AIDS6","AIDS7","AIDS8","AIDS9","AIDS10")
for(i in 1:n)
{
  if(ai_dependency_score[i]==1)
  {
    AI_Dependency_Score[i,]=c(0,0,0,0,0,0,0,0,0)
  }else if(ai_dependency_score[i]!=1)
  {
    if(ai_dependency_score[i]==2)
    {
      AI_Dependency_Score[i,]=c(1,0,0,0,0,0,0,0,0)
    }else if(ai_dependency_score[i]==3 | ai_dependency_score[i]==4 | ai_dependency_score[i]==5)
    {
      if(ai_dependency_score[i]==3)
      {
        AI_Dependency_Score[i,]=c(1,1,0,0,0,0,0,0,0)
      }else if(ai_dependency_score[i]==4)
      {
        AI_Dependency_Score[i,]=c(1,1,1,0,0,0,0,0,0)
      }else
      {
        AI_Dependency_Score[i,]=c(1,1,1,1,0,0,0,0,0)
      }
    }else if(ai_dependency_score[i]==6 | ai_dependency_score[i]==7 | ai_dependency_score[i]==8 | ai_dependency_score[i]==9)
    {
      if(ai_dependency_score[i]==6)
      {
        AI_Dependency_Score[i,]=c(1,1,1,1,1,0,0,0,0)
      }else if(ai_dependency_score[i]==7)
      {
        AI_Dependency_Score[i,]=c(1,1,1,1,1,1,0,0,0)
      }else
      {
        if(ai_dependency_score[i]==8)
        {
          AI_Dependency_Score[i,]=c(1,1,1,1,1,1,1,0,0)
        }else
        {
          AI_Dependency_Score[i,]=c(1,1,1,1,1,1,1,1,0)
        }
      }
    }else
    {
        AI_Dependency_Score[i,]=c(1,1,1,1,1,1,1,1,1)
    }
  }
}
CD5=data.frame(ai_dependency_score,AI_Dependency_Score);CD5
```
### 6.Score as per AI Ethics
```{r}
ai_ethics_score
```
```{r}
AI_Ethics_Score=matrix(0,ncol=9,nrow=n)
colnames(AI_Ethics_Score)=c("AIES2","AIES3","AIES4","AIES5","AIES6","AIES7","AIES8","AIES9","AIES10")
for(i in 1:n)
{
  if(ai_ethics_score[i]==1)
  {
    AI_Ethics_Score[i,]=c(0,0,0,0,0,0,0,0,0)
  }else if(ai_dependency_score[i]!=1)
  {
    if(ai_ethics_score[i]==2)
    {
      AI_Ethics_Score[i,]=c(1,0,0,0,0,0,0,0,0)
    }else if(ai_ethics_score[i]==3 | ai_ethics_score[i]==4 | ai_ethics_score[i]==5)
    {
      if(ai_ethics_score[i]==3)
      {
        AI_Ethics_Score[i,]=c(1,1,0,0,0,0,0,0,0)
      }else if(ai_ethics_score[i]==4)
      {
        AI_Ethics_Score[i,]=c(1,1,1,0,0,0,0,0,0)
      }else
      {
        AI_Ethics_Score[i,]=c(1,1,1,1,0,0,0,0,0)
      }
    }else if(ai_ethics_score[i]==6 | ai_ethics_score[i]==7 | ai_ethics_score[i]==8 | ai_ethics_score[i]==9)
    {
      if(ai_ethics_score[i]==6)
      {
        AI_Ethics_Score[i,]=c(1,1,1,1,1,0,0,0,0)
      }else if(ai_ethics_score[i]==7)
      {
        AI_Ethics_Score[i,]=c(1,1,1,1,1,1,0,0,0)
      }else
      {
        if(ai_ethics_score[i]==8)
        {
          AI_Ethics_Score[i,]=c(1,1,1,1,1,1,1,0,0)
        }else
        {
          AI_Ethics_Score[i,]=c(1,1,1,1,1,1,1,1,0)
        }
      }
    }else
    {
        AI_Ethics_Score[i,]=c(1,1,1,1,1,1,1,1,1)
    }
  }
}
CD6=data.frame(ai_ethics_score,AI_Ethics_Score);CD6
```
### 7.Score as per Concept Understanding
```{r}
concept_understanding_score
```
```{r}
Concept_Understanding_Score=matrix(0,ncol=9,nrow=n)
colnames(Concept_Understanding_Score)=c("CU2","CU3","CU4","CU5","CU6","CU7","CU8","CU9","CU10")
for(i in 1:n)
{
  if(concept_understanding_score[i]==1)
  {
    Concept_Understanding_Score[i,]=c(0,0,0,0,0,0,0,0,0)
  }else if(concept_understanding_score[i]!=1)
  {
    if(concept_understanding_score[i]==2)
    {
      Concept_Understanding_Score[i,]=c(1,0,0,0,0,0,0,0,0)
    }else if(concept_understanding_score[i]==3 | concept_understanding_score[i]==4 | concept_understanding_score[i]==5)
    {
      if(concept_understanding_score[i]==3)
      {
        Concept_Understanding_Score[i,]=c(1,1,0,0,0,0,0,0,0)
      }else if(concept_understanding_score[i]==4)
      {
       Concept_Understanding_Score[i,]=c(1,1,1,0,0,0,0,0,0)
      }else
      {
        Concept_Understanding_Score[i,]=c(1,1,1,1,0,0,0,0,0)
      }
    }else if(concept_understanding_score[i]==6 | concept_understanding_score[i]==7 | concept_understanding_score[i]==8 | concept_understanding_score[i]==9)
    {
      if(concept_understanding_score[i]==6)
      {
        Concept_Understanding_Score[i,]=c(1,1,1,1,1,0,0,0,0)
      }else if(concept_understanding_score[i]==7)
      {
        Concept_Understanding_Score[i,]=c(1,1,1,1,1,1,0,0,0)
      }else
      {
        if(concept_understanding_score[i]==8)
        {
          Concept_Understanding_Score[i,]=c(1,1,1,1,1,1,1,0,0)
        }else
        {
          Concept_Understanding_Score[i,]=c(1,1,1,1,1,1,1,1,0)
        }
      }
    }else
    {
       Concept_Understanding_Score[i,]=c(1,1,1,1,1,1,1,1,1)
    }
  }
}
CD7=data.frame(concept_understanding_score,Concept_Understanding_Score);CD7
```


### 8.Score as per Class Participation
```{r}
class_participation_score
```
```{r}
Class_Participation_Score=matrix(0,ncol=9,nrow=n)
colnames(Class_Participation_Score)=c("CPS2","CPS3","CPS4","CPS5","CPS6","CPS7","CPS8","CPS9","CPS10")
for(i in 1:n)
{
  if(class_participation_score[i]==1)
  {
    Class_Participation_Score[i,]=c(0,0,0,0,0,0,0,0,0)
  }else if(class_participation_score[i]!=1)
  {
    if(class_participation_score[i]==2)
    {
      Class_Participation_Score[i,]=c(1,0,0,0,0,0,0,0,0)
    }else if(class_participation_score[i]==3 | class_participation_score[i]==4 | class_participation_score[i]==5)
    {
      if(class_participation_score[i]==3)
      {
        Class_Participation_Score[i,]=c(1,1,0,0,0,0,0,0,0)
      }else if(class_participation_score[i]==4)
      {
       Class_Participation_Score[i,]=c(1,1,1,0,0,0,0,0,0)
      }else
      {
        Class_Participation_Score[i,]=c(1,1,1,1,0,0,0,0,0)
      }
    }else if(class_participation_score[i]==6 | class_participation_score[i]==7 | class_participation_score[i]==8 | class_participation_score[i]==9)
    {
      if(class_participation_score[i]==6)
      {
        Class_Participation_Score[i,]=c(1,1,1,1,1,0,0,0,0)
      }else if(class_participation_score[i]==7)
      {
        Class_Participation_Score[i,]=c(1,1,1,1,1,1,0,0,0)
      }else
      {
        if(class_participation_score[i]==8)
        {
          Class_Participation_Score[i,]=c(1,1,1,1,1,1,1,0,0)
        }else
        {
          Class_Participation_Score[i,]=c(1,1,1,1,1,1,1,1,0)
        }
      }
    }else
    {
       Class_Participation_Score[i,]=c(1,1,1,1,1,1,1,1,1)
    }
  }
}
CD8=data.frame(class_participation_score,Class_Participation_Score);CD8
```
```{r}
attach(data)
```

```{r}
Data=data.frame(Gender,
                Grade_Level,
                AI_Tools_Used,
                AI_Usage_Purpose,
                AI_Dependency_Score,
                AI_Ethics_Score,
                Concept_Understanding_Score,
                Class_Participation_Score,
                age,
                study_hours_per_day,
                ai_usage_time_minutes,
                ai_generated_content_percentage,
                ai_prompts_per_week,
                last_exam_score,
                assignment_scores_avg,
                attendance_percentage,
                study_consistency_index,
                improvement_rate,
                sleep_hours,
                social_media_hours,
                tutoring_hours,
                final_score,
                passed);Data

```

```{r}
hist(Data$final_score,col="yellow",freq=F,
     main="Histogram of Percentage of Final Score",
     xlab="Percentage of Final Score"
     )
```

```{r}
library(ggplot2)
ggplot(Data, aes(x = passed)) +
  geom_bar(aes(y = (..count..) / sum(..count..)),
           fill = "darkblue") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Relative Frequency Distribution of Pass Status",
       x = "Pass Status",
       y = "Relative Frequency (%)")
```



```{r}
id=1:8000
n_train=0.8*8000;n_train
n_test=0.2*8000;n_test
set.seed(123)
sample_id=sample(id,n_train,replace=F)
training_data=Data[sample_id,]
test_data=Data[-sample_id,]
```
```{r}
ncol(test_data)
```

## Multiple Linear Regression:

### Train_test_split:
```{r}
training_data_mlr=training_data[,-68]
test_data_mlr=test_data[,-68]
test_data_mlr
```

### Fitted mlr model using training data:
```{r}
mlr_train=lm(final_score~.,training_data_mlr)
```

```{r}
AIC(mlr_train)
```
### Summary of fitted mlr model using training data:
```{r}
summary(mlr_train)
library(stargazer)
stargazer(mlr_train,type="text",out="f2.txt")
```

### MSE
```{r}
y_est_train_mlr=predict(mlr_train,training_data_mlr)
y_est_test_mlr=predict(mlr_train,test_data_mlr)
train_MSE_mlr=mean((training_data_mlr$final_score-y_est_train_mlr)^2);train_MSE_mlr
test_MSE_mlr=mean((test_data_mlr$final_score-y_est_test_mlr)^2);test_MSE_mlr
AIC(mlr_train)
```

## Logistic Regression:

### Train_test_split
```{r}
training_data_lr=training_data[,-67]
test_data_lr=test_data[,-67]
test_data_lr
```

### Fitted Logistic Regression model using training data:
```{r}
lr_train=glm(passed~.,training_data_lr,family="binomial")
```

### Summary of the fitted Logistic Regression model using training data:
```{r}
summary(lr_train)
library(stargazer)
stargazer(lr_train,type="text",out="lr.txt")
```

```{r}
prob_est_test=predict(lr_train, newdata = test_data_lr, type = "response")
cut_point=0.5
final_pred=ifelse(prob_est_test> cut_point, 1, 0)
cm=table(Predicted = final_pred, Actual = test_data_lr$passed)
cm
TP=ifelse("1" %in% rownames(cm) & "1" %in% colnames(cm), cm["1","1"], 0)
TN=ifelse("0" %in% rownames(cm) & "0" %in% colnames(cm), cm["0","0"], 0)
FP=ifelse("1" %in% rownames(cm) & "0" %in% colnames(cm), cm["1","0"], 0)
FN=ifelse("0" %in% rownames(cm) & "1" %in% colnames(cm), cm["0","1"], 0)
TPR=ifelse((TP + FN) == 0, 0, TP / (TP + FN));TPR
FPR=ifelse((FP + TN) == 0, 0, FP / (FP + TN));FPR 
Precision=ifelse((TP + FP) == 0, 0, TP / (TP + FP));Precision
Recall=TPR;Recall
```
```{r}
ncol(training_data_lr)
```



